---
title: "Introduction to R and R Studio"
author: "Robin Jeffries"
date: "\ `r Sys.Date()`"
output:
  ioslides_presentation:
  beamer_presentation: default
---

##  Overview
This lesson will provide information on how to

1. Install R  -- _The programming language_
2. Install R Studio -- _A fantastic interface for coding in R_
3. Set up preferences -- _To avoid common mistakes and pitfalls_
4. Install Packages -- _Adds functionality to R_


## Installing R & R Studio

Download the latest versions of R and R Studio for your flavor of operating system.

* R [http://cran.r-project.org/](http://cran.r-project.org/)
* R Studio [http://www.rstudio.com/products/rstudio/download/](http://www.rstudio.com/products/rstudio/download/).

If you are having difficulty installing these programs here are links to helpful YouTube video on installing both R and R Studio. 

- [Mac](https://www.youtube.com/watch?v=cX532N_XLIs)
- [Windows](http://www.youtube.com/watch?v=eD07NznguA4)


## Layout of R Studio

- The panel in the upper right contains your *workspace* (Environment) as well as a 
history of the commands that you've previously entered.  
- Any plots that you generate will show up in the panel in the lower right corner.
- There are multiple tabs down there such as your file explorer, and the help files. 
- The panel on the left is the *console* and it's where the action happens.

## The Console

- Everytime you launch RStudio, it will have the same text at the top of the 
console telling you the version of R that you're running.  
- Below that information is the *prompt*, `>` . As its name suggests, this prompt is really a request, a request for a command. 
- Initially, interacting with R is all about typing commands and interpreting the output. 
- These commands and their syntax have evolved over decades (literally) and now provide what many users feel is a fairly natural way to access data and organize, describe, and invoke statistical computations.


## Initial Setup
To avoid some of the most common errors when learning R let's set some preferences in RStudio. 

1. On your computer, create a folder called `RBootcamp`. Store all files associated with this class here.
2. In RStudio the file menu go to _Tools_ then _Global Options_. 
3. Select your `RBootcamp` folder as your default working directory by clicking [Browse.] and navigating to your `RBootcamp` folder. 
4. Uncheck "Restore .RData into workspace at startup" and "Always save history"
5. Where it says 'Save workspace to .RData on exit:" Select 'Never'
6. Check" "Remove duplicate entries in history"
7. Click _apply_ then _ok_. 

## Installing Packages 
Packages are sets of predefined code, functions and data sets. They provide additional functionality and goodies beyond what is included when you download Base R. 

<div class="red">Packages are only installed once per computer ever</div> 

During this course you will be asked to complete several interactive tutorials that use code found in the `swirl` package. 

To install the Swirl package type the following command at the prompt `>` and hit enter to submit the command.
```{r install_swirl, eval=FALSE}
install.packages("swirl")
```


## Installing Swirl lessons

The `swirl` package and it's fantastic tutorials were created by the folks at [http://swirlstats.com/](http://swirlstats.com/). All of their work is free to use and modify under the GPLv3 copyright license. 

I have spliced and diced the lessons to fit this class. My thanks to them for doing all the hard work. 

To install the custom lessons built specifically for this class first load the library to get access to the `install_course_url` function. Then use it to download and install the file directly from my GitHub repository. 
```{r install_swirl_course, eval=FALSE}
library(swirl)
install_course_url("https://github.com/norcalbiostat/R-Bootcamp/blob/master/R%20Bootcamp.zip")
```

## Using Swirl
That's material for Lab 1. 

Go download it from [here](here) and start working on it. 

Don't forget to use the [Piazza](link to piazza course) if you get stuck.






