---
title: 'Lab 3: Generating and working with Data'
output:
  pdf_document: 
    highlight: zenburn
    number_sections: yes
  html_document:
    highlight: pygments
    theme: cerulean
---

# Introduction 
The goal of this lab is to introduce you to how R deals with the most common way data is stored for analysis, a rectangular format like an Excel spreadsheet. Ideally each row represents data on a single observation and each column contains data on a single variable, or characteristic, of the observation. This is called `tidy data`. We will start to learn some tools to look at the data, and for getting data from an external file into R for analysis. 


## Matricies and Data Frames

Complete the **Matricies and Data Frames** Swirl lesson. 


### Practice
Start a new script file. Recreate the following data frame by typing your commands in the code editor. Save the data frame as an object named `iris2`.  

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr); library(readr)
a <- iris %>%  group_by(Species) %>% slice(1)
knitr::kable(a)
```


## Looking at Data

Complete the **Looking at Data** Swirl lesson. 

### Practice
R comes pre-loaded with some training data sets you can practice with. One of those is the full `iris` data by Edgar Anderson. You can read more about this data set by typing `?iris`

1. Compare your `iris2` to the top 4 rows of the `iris` data using `head()`. 
2. Use `summary()` to find out how many different species of flowers were sampled, and how many flowers from each species were sampled.  


## Tidy Data

Review the lecture notes on [Tidy Data](Link).

The following is data collected by the Pew Research Center that examines the relationship between income and religion in the US. In other words, which religions have the wealthiest adherents? 
```{r, echo=FALSE}
pew <- read_csv("tidy/pew.csv")
knitr::kable(pew[-5,1:8])
```

1. What characteristics are being measured here? Hint: There are 3. 
2. Is this tidy data? 
3. What would a tidy version of this data set look like? 


## Importing Data

In this bootcamp we are only going to explore reading flat files that exist on your computer into R from three most commonly used data sources: A text file, A CSV file and an Excel file. 

Go to the [Course schedule](http://norcalbiostat.github.io/R-Bootcamp/Schedule.html) and download the files for this lab. Be sure to put into your RBootcamp folder. You have to \textcolor{red}{right click} and select \textcolor{red}{save as} to correctly save the Email file}. Otherwise it will open in a browser window and you _cannot_ save it from there correctly. Once they are in your RBootcamp folder, go ahead and open them just to see what they look like. 

#### Text and CSV files
Text files are very simple files that have a `.txt` file extension. Columns are separated by a `delimiter`. Common delimiters include a space, a comma (.) or a tab. Uncommon delimiters could include a % or even a semi-colon. Follow along with these examples and make sure you can read in the data correctly and that it matches what is shown below. 

We will use the `read.table()` function that is in base R to read in any type of delimited file. Don't forget that you need to update the path shown to **YOUR** path on **YOUR** machine that points to your RBootcamp folder.

A tab delimted text files can be read in using `"\t"` as the deliminator character. In this class you **ALWAYS** want to include `header=TRUE` to signifiy that the data in the first row contains our column names. 
```{r}
email50 <- read.table("../data/email50.txt", header=TRUE, sep="\t")
dim(email50)
class(email50)
email50[1:6,1:6]
```

CSV is a fancy way of saying a text file with comma separated values (i.e. CSV). We could use `read.table()` but `read.csv()` is optimized to read in CSV files. 

```{r}
NCbirths <- read.csv("../data/NCbirths.csv", header=TRUE)
dim(NCbirths)
class(NCbirths)
head(NCbirths)
str(NCbirths)
```


#### CSV and Excel files
The best method I have found so far to read in Excel files is from the `readxl` packages by Hadley Wickham. This packages need to be installed first, and then can be simply loaded each time you start an `R` session where you will be reading in this type of data. Go ahead and install it now.  

The `read_excel()` function is what we are going to use. Note the use of the underscore `_` instead of a period `.` between read and the file type.  

```{r, eval=-6}
library(readxl)
OSCounty <- read_excel("../data/OSCounty.xlsx", sheet=1, col_names=TRUE)
dim(OSCounty)
class(OSCounty)
OSCounty[1:6,1:6]
str(OSCounty) #Not run due to the length of output. 
```

Notice that `OSCounty` isn't just a data frame, but also of class `tbl_df`. It's just another format of data storage. You are welcome to look up the differnces and any advantages of using `tbl_df` over `data.frame` on your own time, we will not cover those differences in this class. 





